```{r setup, include=FALSE}
### Load packages
library(tidyverse)
library(ggplot2)
library(jrvFinance)
### Set background and foreground colors
plot_bg <- "#FFFFFF" # "#191919"
plot_fg <- "#000000" # #929292"
plot_fg_alt <- "#969696" # ??
```

# Convexity {#Convexity}

Ok, so now you know how to price bonds. My next big goal is to introduce *convexity* and explain how to think about it. You're going to see what convexity is, why bonds are convex, and how convexity can help you build a good investment portfolio. 

## What is Convexity? 

The mathematical definition of convexity talks about the second derivative of the your investment's payoff function, but the idea is so much simpler than that. All we need is two plots. The first plot shows how the price of a 10 year bond changes as interest rate varies from 0.5% to 4%.  

```{r Bonds-Are-Convex, echo=FALSE, out.width = "80%", fig.align = "center", message = FALSE}
x_min <- 0.5; x_max = 4
yield <- seq(from = x_min, to = x_max, by = 0.01)
price <- 1000 * sapply( yield, function(x) {
  bond.price(settle = "2019-01-01", mature = "2029-01-01",
             coupon = 0.02, yield = x) } )
Price_Yield <- tibble("Yield" = yield, "Price" = price)

ggplot() + 
  geom_line(data = Price_Yield, aes(Yield, Price), linetype = "solid") +
  scale_y_continuous(label = scales::percent) +
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Bond Price vs. Interest Rate",
       x = "Interest Rate",
       y = "Price") +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, face="bold", 
                                  margin = margin(10, 0, 10, 0),
                                  colour = plot_fg),
        text = element_text(size=16), 
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = plot_bg, color = plot_bg),
        plot.background = element_rect(fill = plot_bg, color = plot_bg),
        axis.line = element_line(colour = plot_bg),
        axis.title.x = element_text(margin = unit(c(4, 0, 0, 0), "mm"), 
                                    colour = plot_fg),
        axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm"), 
                                    angle = 90, colour = plot_fg) )
```

The key observation to make is that this curve isn't a straight line. The next plot adds the price of a make-believe asset (which is *not* a bond) which goes up in a straight line. 

```{r Bonds-Are-Convex-2, echo=FALSE, out.width = "80%", fig.align = "center", message = FALSE}
x_min <- 0.5; x_max = 4
yield <- seq(from = x_min, to = x_max, by = 0.01)
price <- 1000 * sapply( yield, function(x) {
  bond.price(settle = "2019-01-01", mature = "2029-01-01",
             coupon = 0.02, yield = x) } ) # also: bond.duration
idx <- 151
m <- ( price[idx] - price[idx+1] ) / ( yield[idx] - yield[idx+1] )
b <- ( price[idx] - m*yield[idx] )
linear = m*yield + b

Price_Yield <- tibble("Yield" = yield, 
                      "Price" = price, 
                      "Linear" = linear)

ggplot() + 
  geom_line(data = Price_Yield, aes(Yield, Price), linetype = "solid") +
  geom_line(data = Price_Yield, aes(Yield, Linear), linetype = "solid") +
  scale_x_continuous(label = scales::percent) +
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Bond Price vs. Yield",
       x = "Yield", y = "Price") +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, face="bold", 
                                  margin = margin(10, 0, 10, 0),
                                  colour = plot_fg),
        text = element_text(size=16), 
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = plot_bg, color = plot_bg),
        plot.background = element_rect(fill = plot_bg, color = plot_bg),
        axis.line = element_line(colour = plot_bg),
        axis.title.x = element_text(margin = unit(c(4, 0, 0, 0), "mm"), 
                                    colour = plot_fg, face = "bold"),
        axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm"), 
                                    angle = 90, colour = plot_fg, face = "bold") )
```

The key observation is that the value of the bond (the curved line) is *always* above the value of the straight line. In other words, the price of a bond goes up *faster* than a straight line does, and goes down *slower*. If we unpack that, it means that when the price of your bond starts going up, it goes up faster and faster. We can see this on the plot because the curve keeps getting steeper as yields go lower. And when the price of your bond goes down, it goes down slower and slower. We can also see this on the plot because the curve flattens out as yields go higher. 

That's *convexity*. It gives a natural asymmetry to your bond investment. You lose money slower and slower as the price goes down, but make it faster and faster as the price goes up. I view convexity as a risk management tool because it naturally reduces our losses *and* increases our gains. This is why convexity is so desirable, and we get it for free with bonds. 


## Why Are Bonds Convex? 

Bonds are *convex* because of compound interest. The thing to compare it against is *simple interest*. We're going start by finding the price of a bond when we get simple interest, and then compare the price/yield chart under simple interest to the one up above that uses compound interest.

A bank account that pays compound interest gives you interest on the interest you earn. We've seen compound interest at work when we looked at the time value of money. We deposited \$100 at 1.5%, and watched it grow to \$101.50 in one year, \$103.02 in two years, and \$104.57 in three years.

  - In the first year, we made 1.5% on \$100, which is \$1.50. 
  - In the second year, we made 1.5% on \$101.50, which is \$1.52. 
  - In the third year, we made 1.5% of \$103.52, which is \$1.55.

Every year, we're making slightly more from the 1.5% because of the interest payments being put into our account. No bank pays simple interest - they all pay compound interest. We're not going to be using it price anything - it's just a tool for us to get some intuition about *why* bonds are convex. 

Under simple interest, the value of our account grows *linearly* over time - it goes up in a straight line. Let's see how our \$100 grows under simple interest. 

  - At the end of the first year, our account is worth \$100 x (1 + 0.015 x 1) = \$101.50. 
  - At the end of the second year, our account is worth \$100 x (1 + 0.015 x 2) = \$103.00.
  - At the end of the third year, our account is worth \$100 x (1 + 0.015 x 3) = \$104.50. 
  
Another way to think of simple interest is that it pays the same amount every year. We're being paid 1.5% on our original deposit of \$100 every year, which is \$1.50. The next thing we want to do is see how the value of these two accounts are affected by the interest rate. I'm going to plot their value after ten years, and vary the interest rate from 0% to 10%. 
 
```{r Simple-Interest-Growth, echo=FALSE, out.width = "80%", fig.align = "center", message = FALSE}
x_min <- 0; x_max = 10
yield <- seq(from = x_min, to = x_max, by = 0.01)
price_compound <- sapply( yield, function(x) { 100 * (1 + x/100)^10 } ) 
price_linear <- sapply( yield, function(x) { 100 * (1 + 10*x/100) } )

Price_Yield <- tibble( "Yield" = yield/100, 
                       "Compound" = price_compound, 
                       "Linear" = price_linear )

ggplot() + 
  geom_line(data = Price_Yield, aes(Yield, Compound, color = plot_fg), 
            linetype = "solid") +
  geom_line(data = Price_Yield, aes(Yield, Linear, color = plot_fg_alt), 
            linetype = "solid") +
  scale_x_continuous(label = scales::percent) + 
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Simple Interest vs. Compound Interest",
       x = "Interest Rate",
       y = "Price") +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, face="bold", 
                                  margin = margin(10, 0, 10, 0),
                                  colour = plot_fg),
        text = element_text(size=16), 
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = plot_bg, color = plot_bg),
        plot.background = element_rect(fill = plot_bg, color = plot_bg),
        axis.line = element_line(colour = plot_bg),
        axis.title.x = element_text(margin = unit(c(4, 0, 0, 0), "mm"), 
                                    colour = plot_fg, face = "bold"),
        axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm"), 
                                    angle = 90, colour = plot_fg, face = "bold") )
```

The account earning simple interest is a straight line, while the account earning compound interest is curved. We've got one more plot to go - this one shows the present value (or *price*) of \$100 in each account 10 years in the future, under compound interest and simple interest. As before, the interest rate varies from 0% to 10%.

```{r Simple-Interest-Price, echo=FALSE, out.width = "80%", fig.align = "center", message = FALSE}
x_min <- 0; x_max = 10
yield <- seq(from = x_min, to = x_max, by = 0.01)
price_compound <- sapply( yield, function(x) { 100 * (1 + x/100)^-10 } ) 
price_linear <- sapply( yield, function(x) { 100 / (1 + 10*x/100) } )

Price_Yield <- tibble( "Yield" = yield/100, 
                       "Compound" = price_compound, 
                       "Linear" = price_linear )

ggplot() + 
  geom_line(data = Price_Yield, aes(Yield, Compound, color = plot_fg), 
            linetype = "solid") +
  geom_line(data = Price_Yield, aes(Yield, Linear, color = plot_fg_alt), 
            linetype = "solid") +
  scale_x_continuous(label = scales::percent) + 
  scale_y_continuous(label = scales::dollar) +
  labs(title = "Simple Interest vs. Compound Interest",
       x = "Interest Rate",
       y = "Price") +
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, face="bold", 
                                  margin = margin(10, 0, 10, 0),
                                  colour = plot_fg),
        text = element_text(size=16), 
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = plot_bg, color = plot_bg),
        plot.background = element_rect(fill = plot_bg, color = plot_bg),
        axis.line = element_line(colour = plot_bg),
        axis.title.x = element_text(margin = unit(c(4, 0, 0, 0), "mm"), 
                                    colour = plot_fg, face = "bold"),
        axis.title.y = element_text(margin = unit(c(0, 4, 0, 0), "mm"), 
                                    angle = 90, colour = plot_fg, face = "bold") )
```
