--- 
title: "Understanding Bonds"
subtitle: "" 
author: "Kamal Rai"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: ""
---

# Introduction {-}

## Preface {-}

Bonds are massively underappreciated by the people I know as an investment option. Maybe it's because bonds never get in the headlines like tech stocks or the S&P 500 do, or because they think bonds are "boring". I haven't been able to figure out why, but I'd like to bring excitement back to bonds. To me, their risk-reward characteristics make them the more important part of anyone's investment portfolio. 

The biggest problem with money is that it causes too much stress. I think the biggest reason for this is that people don't know how to set and pursue financial goals. One part of that is budgeting and saving money. But I'd bet you're very good at understanding how much money you're making and how much you can spend. You may want to spend more (really badly!), but you know what you can afford to buy. And if you buy something that's too expensive or a "waste of money", you'll probably realize it soon enough. So while you might have trouble meeting savings goals, you probably understand the ideas behind them almost intutitively. 

It makes sense, too, because making money and spending it is part of your everyday life. What you don't have as much experience with or intuition about is investing. You don't go around staring at stocks and bonds like you do jeans or computers, trying to figure out which one you should buy. Investors on Wall Street might, but the rest of us have better things to do.

You'll have heard some advice along the way - start saving when you're young, invest in stocks for the long run, buy some bonds if you're a conservative investor - and it might work for you. But it's one-size-fits-all advice, and that limits how well it fits your goals and your tolerance for stress from your investments. It puts your savings at the mercy of a cookie-cutter plan, and you can do better than that. 

If you understand bonds, you can use them to build an investment portfolio that suits you. You have to have set goals first, because your portfolio should be designed to help you reach them. But once you have goals and have constructed a portfolio to pursue them, maintaining it doesn't have to take much time. If you want, you can build portfolios that are reasonably recession-proof and generally only needs to be reviewed monthly or quarterly. 

If you want to take more control of you financial future, this book might be for you. I'm focusing on bonds because the ideas behind thoughtfully investing in bonds are fundamentally important. You'll be able to use them to  and bonds themselves have lots of attractive properties. The  

## Acknowledgments {-}

This book is written in Rmarkdown using the *bookdown* package, hosted on Netlify, and built using Travis CI. The *Authoring Books and Technical Documents with R Markdown* at https://musing-aryabhata-b16338.netlify.com/ was very helpful in setting up Netlify and Travis CI. 

```{r, results='asis', echo = FALSE, eval = TRUE} 
# if(knitr:::pandoc_to() == "html") { 
#   cat('## Sign Up For Updates {-} \n Click below to sign up to receive notifications every time new chapters # are published \n.') }
# if(knitr:::pandoc_to() == "html") { 
#   cat('<script async data-uid="244f107410"src="https://hustling-trailblazer-1491.ck.page/244f107410/index.js"># </script>') }
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
